A Classical Density-Functional Theory for Describing Water Interfaces

### 2012-12-10 papers/water-SAFT/paper.tex
changes to water paper Tests skipped!

A number of classical density functionals have been developed for
water~\cite{ding1987, Yang1992, yang1994density, gloor2002saft,
  gloor2004accurate, gloor2007prediction, Jaqaman2004,
  clark2006developing, chuev2006, lischner2010classical,
  fu2005vapor-liquid-dft,kiselev2006new, blas2001examination}, each of
which captures some of the qualitative behavior of water.  However,
each of these functionals also fail to capture some of water's unique
properties.  For instance, the functional of Lischner \emph{et
  al}\cite{lischner2010classical} treats the surface tension
correctly, but can only be used at room temperature, and thus captures
none of the temperature-dependence of water.  A functional by Chuev
and Skolov\cite{chuev2006} uses an ad hoc modification of FMT that can
predict hydrophobic hydration near temperatures of 298~K, but does not
produce a correct equation of state due to their method producting a
high value for pressure. A number of classical density functionals
have recently been produced that are based on Statistical Associating
Fluid Theory (SAFT)\cite{ segura1997associating, segura1998comparison,
  yu2002fmt-dft-inhomogeneous-associating,
  fu2005vapor-liquid-dft,gloor2002saft,muller2001molecular,
  clark2006developing, gloor2007prediction, gloor2004accurate,
  gross2009density, kahl2008modified, blas2001examination}.  These
functionals are based on a perturbative thermodynamic expansion, and
do reproduce the temperature-dependence of water's properties.

@@

Functionals constructed using a local density approximation
fail to satisfy the contact-value theorem, and
therefore incorrectly model small hard solutes.
The contact-value theorem states that the pressure any fluid exerts
on a hard wall interface is proportional to the contact density of the
fluid\cite{henderson1979exact}:
\begin{equation}\label{eq:contact}
  p = n_ck_BT,
\end{equation}
where $n_c$ is the contact density, $k_B$ is the Boltzmann constant
and $T$ is the temperature of
the fluid. This leads to the property that the excess chemical
potential of a small hard solute is proportional to the
solvent-excluded volume:
\begin{equation}\label{contactvaluethm}
  F = n k_BT V.
\end{equation}
The contact-value theorem is satisfied by classical density
functionals in which the only purely local term is the ideal gas
contribution to the free energy, and conversely, this theorem is not
satisfied by functionals built using a local density approximation.

@@

The majority of the empirical parameters used in our functional are
taken from the paper of Clark \emph{et al} on developing an optimal
SAFT model for water\cite{clark2006developing}.  This SAFT model
contains five empirical parameters: the hard-sphere radius, an energy
and length scale for the dispersion interaction, and an energy and
length scale for the association interaction.  In addition to the five
empirical parameters of Clark \emph{et al}, we add a single additional
dimensionless parameter $\lscale$---with a fitted value of
0.353---which determines the length scale over which the density is
averaged when computing the dispersion free energy and its derivative.
We determine this final parameter by fitting the computed surface
tension to the experimental surface tension with the result shown in
Figure~\ref{fig:surface-tension}.  Because the SAFT model of Clark
\emph{et al} overestimates the critical temperature---which is a
common feature of SAFT-based functionals that do not explicitly treat
the critical point---we cannot reasonably describe the surface tension
at all temperatures, and choose to fit the surface tension at and
around room temperature.

@@

We begin by studying a single hydrophobic rod immersed in water.
% We
% model the density on a grid with spacing of about 0.01~nm, which gives
% the free energy of hydration to a precision of better than a percent,
% and the density at contact to a few percent precision.
In Figure~\ref{fig:energy-vs-diameter} we show the excess chemical
potential at room temperature, scaled by the solvent-accessible
surface area of the hard rod, plotted as a function of hard-rod
radius.  We define the hard-rods radius as the radius from which water
is excluded.
For rods with radius larger than 0.5~nm or so, this reaches a maximum value of
75~mN/m, which is slightly higher than the bulk surface.  In the limit
of very large rods, this value will decrease and approach the bulk
surface value.  As seen in the inset of
Fig.~\ref{fig:energy-vs-diameter}, for rods with very small radius (less than
about 0.5~\AA) the excess chemical potential is proportional to volume,
as required by the contact-value theorem (see Equation~\ref{contactvaluethm}).

@@

Starting from the surface energy term, we can calculate the 
free energy per length, which is equal to the circumference multiplied 
by the surface tension. The force per length is the derivative of this
with respect to the separation. 
The circumference of the stadium-shape is
\begin{equation}
C_{s} = 2\pi r +4r+2d
\end{equation}
and so the force per length is equal to twice the surface tension.

@@

Walther \emph{et al}\cite{walther2004hydrophobic} studied the
interactions between two carbon nanotubes, which are geometrically
similar to our hydrophobic rods, using molecular dynamics with the SPC
model for water, and a Lennard-Jones potential for the interaction of
carbon with water, for nanotubes of diameter 1.25~nm and separations
ranging from about 0.3~nm to 1.5~nm.  The SPC model underestimates the
surface tension of water by about 24\%\cite{vega2007surface}, so we
cannot expect this work to provide quantitative agreement with real
water.  Walther \emph{et al} observe a drying transition between the
two nanotubes, which occurs at a smaller radius than our results
suggest.  However, when Walther \emph{et al} disable the attractive
interaction between nanotube and water, the drying effect occurs at
much longer range, in agreement with our results.

@@

\begin{figure}
\begin{center}
\includegraphics[width=\columnwidth]{figs/sphere-energy-vs-diameter}
\end{center}
\caption{ Excess chemical potential per area versus radius for
  a single hydrophobic sphere
  immersed in water. This should have an asymptote equal to the
  surface tension at room temperature, and it agrees well with the
  surface tension in Figure~\ref{fig:surface-tension}. Results from a
  simulation of SPC/E water~\cite{huang2001shs} are shown as circles.
  The horizontal lines show the experimental and SPC/E bulk surface
  tension for water at standard atmospheric temperature and
  pressure. }
\label{fig:sphere-energy-vs-diameter}
\end{figure}

A common model of hydrophobic solutes is the hard-sphere solute, which
is the simplest possible solute, and serves as a test case for
understanding of hydrophobic solutes in
water\cite{sedlmeier2011entropy}.  As in the single rod, we begin by
examining the ratio of the excess chemical potential of the cavity
system to the solvent-accessible surface area
(Figure~\ref{fig:sphere-energy-vs-diameter}).  This
effective surface tension surpasses the bulk surface tension at a
radius of almost 1~nm, and at large radius will drop to the bulk value.  As
with the single rod, we see the analytically correct behavior in the
limit of small solutes.  For comparison, we plot the free energy
calculated using a molecular dynamics simulation of SPC/E
water\cite{huang2001shs}.  The agreement is quite good, apart from the
issue that the SPC/E model for water significantly underestimates the
surface bulk tension of water at room
temperature\cite{vega2007surface}.

>>>>

A number of classical density functionals have been developed for
water~\cite{ding1987, Yang1992, yang1994density, gloor2002saft,
  gloor2004accurate, gloor2007prediction, Jaqaman2004,
  clark2006developing, chuev2006, lischner2010classical,
  fu2005vapor-liquid-dft,kiselev2006new, blas2001examination,
  sundararaman2012computationally}, each of
which captures some of the qualitative behavior of water.  However,
each of these functionals also fail to capture some of water's unique
properties.  For instance, the functional of Lischner \emph{et
  al}\cite{lischner2010classical} treats the surface tension
correctly, but can only be used at room temperature, and thus captures
none of the temperature-dependence of water.  A functional by Chuev
and Skolov\cite{chuev2006} uses an ad hoc modification of FMT that can
predict hydrophobic hydration near temperatures of 298~K, but does not
produce a correct equation of state due to their method producing a
high value for pressure.
A number of classical density functionals
have recently been produced that are based on Statistical Associating
Fluid Theory (SAFT)\cite{ segura1997associating, segura1998comparison,
  yu2002fmt-dft-inhomogeneous-associating,
  fu2005vapor-liquid-dft,gloor2002saft,muller2001molecular,
  clark2006developing, gloor2007prediction, gloor2004accurate,
  gross2009density, kahl2008modified, blas2001examination}.  These
functionals are based on a perturbative thermodynamic expansion, and
do reproduce the temperature-dependence of water's properties.
%
We should give special mention to Sundararaman \emph{et al} recently
introduced a classical density functional for water using an model in
which a water molecule is seen as a hard sphere attached to two
tetrahedrally oriented hard spheres representing voids, or
orientations in which a hydrogen bond may not be formed, with all
attractive interactions being lumped into a single pair potential
treated in a mean field
approximation\cite{sundararaman2012computationally}.

@@

Functionals constructed using a local density approximation fail to
satisfy the contact-value theorem, and therefore incorrectly model
small hard solutes.  The contact-value theorem relates the pressure on
a hard wall interface to the contact density of the fluid at that
interface:
\begin{equation}\label{eq:contact}
  p = n_ck_BT,
\end{equation}
where $n_c$ is the contact density, $k_B$ is the Boltzmann constant
and $T$ is the temperature of the fluid.  The pressure in this formula
is defined as the force per unit \emph{solvent accessible area}, which
differs from the actual pressure when the surface is curved.  For a
cavity in which the solvent is excluded from an arbitrarily small
volume, the contact density will be the same as the bulk density, and
therefore we can integrate the above pressure to find that the excess
chemical potential of a small hard solute is proportional to the
solvent-excluded volume:
\begin{equation}\label{contactvaluethm}
  F = n k_BT V.
\end{equation}
The contact-value theorem is violated by local classical density
functionals such as those using a local density approximation or a
square-gradient term, but is satisfied by non-local classical density
functionals, such as those using a weighted-density approach.

@@

The majority of the empirical parameters used in our functional are
taken from the paper of Clark \emph{et al} on developing an optimal
SAFT model for water\cite{clark2006developing}.  This SAFT model
contains five empirical parameters: the hard-sphere radius, an energy
and length scale for the dispersion interaction, and an energy and
length scale for the association interaction.  In addition to the five
empirical parameters of Clark \emph{et al}, we add a single additional
dimensionless parameter $\lscale$---with a fitted value of
0.353---which determines the length scale over which the density is
averaged when computing the dispersion free energy and its derivative.
We determine this final parameter by fitting the computed surface
tension to the experimental surface tension with the result shown in
Figure~\ref{fig:surface-tension}.  Because the SAFT model of Clark
\emph{et al} overestimates the critical temperature---which is a
common feature of SAFT-based functionals that do not explicitly treat
the critical point---we cannot reasonably describe the surface tension
at all temperatures, and choose to fit the surface tension at and
around room temperature.  We note here that although we could have
chosen to fit the surface tension with a square-gradient term in the
free energy, instead of adjusting the length scale for the dispersive
attraction.  However, would have resulted in a functional that
violates the contact-value theorem, which among other problems would
fail to satisfy Equation~\ref{contactvaluethm} for the excess chemical
potential of small solutes.

@@

We begin by studying a single hydrophobic rod immersed in water.
% We
% model the density on a grid with spacing of about 0.01~nm, which gives
% the free energy of hydration to a precision of better than a percent,
% and the density at contact to a few percent precision.
In Figure~\ref{fig:energy-vs-diameter} we show the excess chemical
potential at room temperature, scaled by the solvent-accessible
surface area of the hard rod, plotted as a function of hard-rod
radius.  We define the hard-rods radius as the radius from which water
is excluded.  For rods with radius larger than 0.5~nm or so, this
reaches a maximum value of 75~mN/m, which is slightly higher than
macroscopic surface tension.  In the limit of very large rods, this
value will decrease and approach the macroscopic value.  As seen in
the inset of Fig.~\ref{fig:energy-vs-diameter}, for rods with very
small radius (less than about 0.5~\AA) the excess chemical potential
is proportional to volume, satisfying Equation~\ref{contactvaluethm},
which results from the contact-value theorem.

@@

Starting from the surface energy term, we can calculate the 
free energy per length, which is equal to the circumference multiplied 
by the surface tension. 
Working out the circumference of the stadium-shape leads us to
\begin{equation}
F = (2\pi r +4r+2d)\gamma
\end{equation}
where $\gamma$ is the surface tension, $r$ is the radius of the rods,
and $d$ is the separation between rods illustrated in
Figure~\ref{fig:density-rods}.  The force per length is the derivative
of the free energy with respect to the separation $d$, from which we
conclude that the force per length is twice the
surface tension.

@@

Walther \emph{et al}\cite{walther2004hydrophobic} studied the
interactions between two carbon nanotubes, which are geometrically
similar to our hydrophobic rods, using molecular dynamics with the SPC
model for water, for nanotubes of diameter 1.25~nm and separations
ranging from about 0.3~nm to 1.5~nm.  In agreement with our findings
for two purely hydrophobic rods, Walther \emph{et al} find that in the
absence of Lennard-Jones attraction between carbon and oxygen, there
is a drying transition at a distance comparable to the diameter of the
nanotube.  In contrast to this, when the attraction between nanotubes
and water is turned on, they find that the drying transition occurs at
much shorter distances, comparable to the diameter of water.

@@

\begin{figure}
\begin{center}
\includegraphics[width=\columnwidth]{figs/sphere-energy-vs-diameter}
\end{center}
\caption{ Excess chemical potential per area versus radius for
  a single hydrophobic sphere
  immersed in water. This should have an asymptote equal to the
  surface tension at room temperature, and it agrees well with the
  surface tension in Figure~\ref{fig:surface-tension}. Results from a
  simulation of SPC/E water~\cite{huang2001shs} are shown as circles.
  The horizontal lines show the experimental and SPC/E macroscopic surface
  tension for water at standard atmospheric temperature and
  pressure. }
\label{fig:sphere-energy-vs-diameter}
\end{figure}

A common model of hydrophobic solutes is the hard-sphere solute, which
is the simplest possible solute, and serves as a test case for
understanding of hydrophobic solutes in
water\cite{sedlmeier2011entropy}.  As in the single rod, we begin by
examining the ratio of the excess chemical potential of the cavity
system to the solvent-accessible surface area
(Figure~\ref{fig:sphere-energy-vs-diameter}).  This
effective surface tension surpasses the macroscopic surface tension at a
radius of almost 1~nm, and at large radius will drop to the macroscopic value.  As
with the single rod, we see the analytically correct behavior in the
limit of small solutes.  For comparison, we plot the free energy
calculated using a molecular dynamics simulation of SPC/E
water\cite{huang2001shs}.  The agreement is quite good, apart from the
issue that the SPC/E model for water significantly underestimates the
macroscopic surface tension of water at room
temperature\cite{vega2007surface}.
